---
title: "Model"
author: "Caleb Stevens cstevens5@wisc.edu"
output: html_document
---

### Setup

**Fill in "this.wd" variable below with the path to your local folder "wilt"** then all chunks should be able to run without further changes

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## SET PATH
this.wd <- "E:/wilt/"
##

# load packages
library(tidyverse)
library(sf)
library(assertthat)
library(mgcv)
library(gridExtra)
library(spatstat)

# construct relative paths
rawdat.wd <- paste0(this.wd, "raw_data/")
cleandat.wd <- paste0(this.wd, "clean_data/")
mod.res.path <- paste0(this.wd, "models/model_obj/")
fun.path <- paste0(this.wd, "code/functions/")
```

# HP3-GAM

## Fit Models

### Load & check data

```{r}
# load data - loads as oak.dat.scale
load(paste0(cleandat.wd, "model_data/oak_dat_scale.Rds"))
```

### Null models

```{r}
#
# H_0A : Oak Wilt = function of spatial dependence + oak population
#
# REML
mod.h0A.reml <- bam(wilt_ct ~ s(x, y) + s(oak_prob, bs="cs"), data=oak.dat.scale, family=poisson(link="log"), method="fREML")
save(mod.h0A.reml, file=paste0(mod.res.path, "mod_h0A_reml.Rds"))
rm(mod.h0A.reml)
# GACV
mod.h0A.gacv <- bam(wilt_ct ~  s(x, y) + s(oak_prob, bs="cs"), data=oak.dat.scale, family=poisson(link="log"), method="GACV.Cp")
save(mod.h0A.gacv, file=paste0(mod.res.path, "mod_h0A_gacv.Rds"))
rm(mod.h0A.gacv)

#
# H_0B : Oak Wilt = function of oak population + environmental factors 
#
# REML - full
mod.h0B.reml.full <- bam(wilt_ct ~ s(oak_prob, bs="cs") + s(elev, bs="cs") + s(bd, bs="cs") + s(ph, bs="cs") + s(sand, bs="cs") + s(ch_net_dist, bs="cs") + s(hs, bs="cr") + s(slope, bs="cr") + s(aspect, bs="cr") + s(plan_curv, bs="cr"), data=oak.dat.scale, family=poisson(link="log"), method="fREML")
save(mod.h0B.reml.full, file=paste0(mod.res.path, "mod_h0B_reml_full.Rds")) 
# summary(mod.h0B.reml.full)
rm(mod.h0B.reml.full)
# REML - subset
mod.h0B.reml.subs <- bam(wilt_ct ~ s(oak_prob, bs="cs") + s(elev, bs="cs") + s(bd, bs="cs") + s(ph, bs="cs") + s(sand, bs="cs") + s(ch_net_dist, bs="cs") + s(hs, bs="cr") + s(aspect, bs="cr"), data=oak.dat.scale, family=poisson(link="log"), method="fREML") # 74963 - slope and plan curv only vars removed
summary(mod.h0B.reml.subs)
save(mod.h0B.reml.subs, file=paste0(mod.res.path, "mod_h0B_reml_subs.Rds"))
rm(mod.h0B.reml.subs)

# GACV - full
mod.h0B.gacv.full <- bam(wilt_ct ~ s(oak_prob, bs="cs") + s(elev, bs="cs") + s(bd, bs="cs") + s(ph, bs="cs") + s(sand, bs="cs") + s(ch_net_dist, bs="cs") + s(hs, bs="cr") + s(slope, bs="cr") + s(aspect, bs="cr") + s(plan_curv, bs="cr"), data=oak.dat.scale, family=poisson(link="log"), method="GACV.Cp")
summary(mod.h0B.gacv.full)
save(mod.h0B.gacv.full, file=paste0(mod.res.path, "mod_h0B_gacv_full.Rds"))
rm(mod.h0B.gacv.full)
# GACV - subs
mod.h0B.gacv.subs <- bam(wilt_ct ~ s(oak_prob, bs="cs") + s(elev, bs="cs") + s(bd, bs="cs") + s(ph, bs="cs") + s(sand, bs="cs") + s(ch_net_dist, bs="cs") + s(hs, bs="cr") + s(aspect) + s(plan_curv, bs="cr"), data=oak.dat.scale, family=poisson(link="log"), method="GACV.Cp") # 0.39471 - slope only var removed
# summary(mod.h0B.gacv.subs)
save(mod.h0B.gacv.subs, file=paste0(mod.res.path, "mod_h0B_gacv_subs.Rds"))
rm(mod.h0B.gacv.subs)
```

### Fit alternative models

```{r}
#
# H_0A : Oak Wilt = function of spatial dependence, oak population, environmental factors
#
# REML - full
mod.alt.reml.full <- bam(wilt_ct ~ s(x, y) + s(oak_prob, bs="cs") + s(elev, bs="cs") + s(bd, bs="cs") + s(ph, bs="cs") + s(sand, bs="cs") + s(ch_net_dist, bs="cs") + s(hs, bs="cr") + s(slope, bs="cr") + s(aspect, bs="cr") + s(plan_curv, bs="cr"), data=oak.dat.scale, family=poisson(link="log"), method="fREML") 

summary(mod.alt.reml.full)
save(mod.alt.reml.full, file=paste0(mod.res.path, "mod_alt_reml_full.Rds"))
rm(mod.alt.reml.full)

mod.alt.reml.full.2 <- bam(wilt_ct ~ s(x, y) + s(oak_prob, bs="cs") + s(elev, bs="cs") + s(bd, bs="cs") + s(ph, bs="cs") + s(silt, bs="cs") + s(rsp, bs="cs") + s(hs, bs="cr") + s(ls, bs="cr") + s(aspect, bs="cr") + s(plan_curv, bs="cr"), data=oak.dat.scale, family=poisson(link="log"), method="fREML") 
summary(mod.alt.reml.full.2)
save(mod.alt.reml.full.2, file=paste0(mod.res.path, "mod_alt_reml_full_2.Rds"))
rm(mod.alt.reml.full.2)

# REML subs

mod.alt.reml.subs <- bam(wilt_ct ~ s(x, y) + s(oak_prob, bs="cs") + s(elev, bs="cs") + s(ph, bs="cs") + s(bd, bs="cs") + s(ch_net_dist, bs="cs") + s(hs, bs="cr") + s(aspect, bs="cr"), data=oak.dat.scale, family=poisson(link="log"), method="fREML")
summary(mod.alt.reml.subs)
# 72620 -> remove slope
# 72733 -> remove plan curve
# 72816 -> remove sand
# 72822 -> remove bd
# 72788 -> add it back
summary(mod.alt.reml.subs)
save(mod.alt.reml.subs, file=paste0(mod.res.path, "mod_alt_reml_subs.Rds"))
rm(mod.alt.reml.subs)
# load(paste0(mod.res.path, "mod_alt_reml_subs.Rds"))

mod.713.reml.tp <- bam(wilt_ct ~ s(x,y) + s(elev, aspect) + s(hs, bs="cs") + s(ch_net_dist, bs="cs") + s(oak_prob, bs="cs") + s(ph, bs="cs") + s(bd, bs="cs"), data=oak.dat.scale, family=poisson(link="log"), method="fREML")
save(mod.713.reml.tp, file=paste0(mod.res.path, "mod_713_reml_tp.Rds"))

# GACV - full
mod.alt.gacv.full <- bam(wilt_ct ~ s(x, y) + s(oak_prob, bs="cs") + s(elev, bs="cs") + s(bd, bs="cs") + s(ph, bs="cs") + s(sand, bs="cs") + s(ch_net_dist, bs="cs") + s(hs, bs="cr") + s(slope, bs="cr") + s(aspect, bs="cr") + s(plan_curv, bs="cr"), data=oak.dat.scale, family=poisson(link="log"), method="GACV.Cp")
summary(mod.alt.gacv.full)
save(mod.alt.gacv.full, file=paste0(mod.res.path, "mod_alt_gacv_full.Rds"))
rm(mod.alt.gacv.full)
# GACV - subs
mod.alt.gacv.subs <- bam(wilt_ct ~ s(x, y) + s(oak_prob, bs="cs") + s(elev, bs="cs") + s(bd, bs="cs") + s(ph, bs="cs") + s(sand, bs="cs") + s(ch_net_dist, bs="cs") + s(hs, bs="cr") + s(plan_curv, bs="cr"), data=oak.dat.scale, family=poisson(link="log"), method="GACV.Cp")
# 0.40609, removed aspect
# 0.4226 removed slope
# 0.447 removed 
summary(mod.alt.gacv.subs)
save(mod.alt.gacv.subs, file=paste0(mod.res.path, "mod_alt_gacv_subs.Rds"))
rm(mod.alt.gacv.subs)
```

### Fit training models for later model comparison

```{r}
# H_A : spatial location, population, env. factors
mod.train.alt.reml <- bam(wilt_train ~ s(x, y) + s(oak_prob, bs="cs") + s(elev, bs="cs") + s(bd, bs="cs") + s(ph, bs="cs") + s(ch_net_dist, bs="cs") + s(hs, bs="cr") + s(aspect, bs="cs"), data=oak.dat.scale, family=poisson(link="log"), method="fREML") 
save(mod.train.alt.reml, file=paste0(mod.res.path, "mod_train_alt_reml.Rds"))
rm(mod.train.alt.reml)

mod.train.alt.gacv <- bam(wilt_train ~ s(x, y) + s(oak_prob, bs="cs") + s(elev, bs="cs") + s(ch_net_dist, bs="cs") + s(hs, bs="cr"), data=oak.dat.scale, family=poisson(link="log"), method="GACV.Cp") 
# 0.38945 -> remove bd and ph, 0.5342
save(mod.train.alt.gacv, file=paste0(mod.res.path, "mod_train_alt_gacv.Rds"))
rm(mod.train.alt.gacv)

# H_0A : spatial dependence + oak population
mod.train.h0A.reml <- bam(wilt_train ~ s(x, y) + s(oak_prob, bs="cs"), data=oak.dat.scale, family=poisson(link="log"), method="fREML")
save(mod.train.h0A.reml, file=paste0(mod.res.path, "mod_train_h0A_reml.Rds"))
rm(mod.train.h0A.reml)

mod.train.h0A.gacv <- bam(wilt_train ~ s(x, y) + s(oak_prob, bs="cs"), data=oak.dat.scale, family=poisson(link="log"), method="GACV.Cp")
save(mod.train.h0A.gacv, file=paste0(mod.res.path, "mod_train_h0A_gacv.Rds"))
rm(mod.train.h0A.gacv)

# H_0B : oak population + terrain & soil (no spatial dependence)
mod.train.h0B.reml <- bam(wilt_train ~ s(oak_prob, bs="cs") + s(elev, bs="cs") + s(bd, bs="cs") + s(ph, bs="cs") + s(ch_net_dist, bs="cs") + s(hs, bs="cr") + s(aspect, bs="cs"), data=oak.dat.scale, family=poisson(link="log"), method="fREML")
save(mod.train.h0B.reml, file=paste0(mod.res.path, "mod_train_h0B_reml.Rds"))
rm(mod.train.h0B.reml)

mod.train.h0B.gacv <- bam(wilt_train ~ s(oak_prob, bs="cs") + s(elev, bs="cs") + s(ph, bs="cs") + s(ch_net_dist, bs="cs") + s(hs, bs="cr"), data=oak.dat.scale, family=poisson(link="log"), method="GACV.Cp")
save(mod.train.h0B.gacv, file=paste0(mod.res.path, "mod_train_h0B_gacv.Rds"))
rm(mod.train.h0B.gacv)

```


## Compare Models 

Comparisons:

 1. $H_{0A}$ vs $H_A$ (REML)
 
 2. $H_{0A}$ vs $H_A$ (GCV)
 
 3. $H_{0B}$ vs $H_A$ (REML)
 
 4. $H_{0B}$ vs $H_A$ (GACV)

```{r}
#' tests hypothesis that the null model actually has the same deviance (is as good as) the alternative model
#'
#' if the model with more predictors (alternative model) fits better than the null model, then the deviance reduction will be large (relative to DOF)
#'
# gam_f_test <- function(mgcv_null_mod, mgcv_alt_mod, DEBUG=F) {
#   delta.dev <- (mgcv_null_mod$deviance - mgcv_alt_mod$deviance)
#   if(DEBUG) {cat(paste0("\nalternative model deviance is lower by ", delta.dev))}
#   assert_that(delta.dev > 0)
#   delta.dof <- sum(mgcv_alt_mod$edf) - sum(mgcv_null_mod$edf)
#   if(DEBUG) {cat(paste0("\nalternative model has additional dof = ", dof))}
#   assert_that(dof > 0)
#   p.val <- pchisq(q=t.stat, df=dof, lower.tail=F)
#   if(DEBUG) {cat(paste0("\np = ", p.val))}
#   if(DEBUG & p.val < 0.01) {cat(paste0("\nthe alternative model appears to explain more of the variation in the data (improved the deviance significantly)"))}
#   if(DEBUG & p.val >= 0.01 & p.val < 0.05) {cat(paste0("\nthe alternative model appears to explain more of the variation in the data (improved the deviance moderately)"))}
#   if(DEBUG & p.val >= 0.05) {cat(paste0("\nno significant evidence that alternative model explains observed data better than the null model"))}
#   return(p.val)
# }

# 1 H_0A vs HA - REML
load(paste0(mod.res.path, "mod_h0A_reml.Rds")) # mod.h0A.reml
load(paste0(mod.res.path, "mod_alt_reml_subs.Rds")) # mod.alt.reml.subs
anova.gam(mod.h0A.reml, mod.alt.reml.subs, test="Chisq")

# 2
load(paste0(mod.res.path, "mod_h0A_gacv.Rds")) # mod.h0A.gacv
load(paste0(mod.res.path, "mod_alt_gacv_subs.Rds"))
anova(mod.h0A.gacv, mod.alt.gacv.subs, test="Chisq")

# 3 - H_{0B} vs H_A REML
load(paste0(mod.res.path, "mod_h0B_reml_full.Rds")) # mod.h0B.reml.full
load(paste0(mod.res.path, "mod_h0B_reml_subs.Rds"))

anova(mod.h0B.reml.subs, mod.alt.reml.subs, test="Chisq")

assert_that(sum(mod.alt.reml.full$edf) >= sum(mod.h0B.reml.full$edf))
p3 <- chi_sq_dev_test(mod.h0B.reml.full, mod.alt.reml.full, T)
rm(mod.h0B.reml.full); rm(mod.alt.reml.full)

# 4 - H_{0B} vs H_A GACV
load(paste0(mod.res.path, "mod_h0B_gacv_subs.Rds")) # mod.h0B.gacv.full
load(paste0(mod.res.path, "mod_alt_gacv_subs.Rds")) # mod.alt.gacv.full
anova(mod.h0B.gacv.subs, mod.alt.gacv.subs, test="Chisq")
```


## Model Results

```{r}
library(mgcViz)

load(paste0(mod.res.path, "mod_alt_gacv_subs.Rds"))
load(paste0(cleandat.wd, "model_data/oak_dat_scale.Rds"))
load(paste0(mod.res.path, "mod_713_reml_tp.Rds"))

b <- getViz(mod.713.reml.tp)

p1 <- plot( sm(b, 1))#; p1
p2 <-  plot( sm(b, 2))#; p2
p3 <-  plot( sm(b, 3))#; p3
p4 <-  plot( sm(b, 4))#; p4
p5 <-  plot( sm(b, 5))#; p5
p6 <-  plot( sm(b, 6))#; p6


p01 <- plot(sm(b, 7)) + labs(title="Elevation Smooth") + theme(title = element_text(size=24, hjust=0.5))

pts.test <- st_read(paste0(cleandat.wd, "sa_points/sa_points.shp")) %>% filter(!is.na(year), year>=2019)
pred <- predict(mod.alt.gacv.subs) %>% 
  as_tibble()
pred$x <- oak.dat.scale$x; pred$y <- oak.dat.scale$y

pred.plot <- pred %>% 
  filter(value>-10, value<0) %>% 
  mutate(relative_risk = 100*((value-min(value))/(max(value)-min(value)))) %>% 
  mutate(x=(x*attr(x, "scaled:scale"))+attr(x, "scaled:center"),
         y=(y*attr(y, "scaled:scale"))+attr(y, "scaled:center"))

p2 <- ggplot() + 
  geom_tile(data=pred.plot, aes(x=x,y=y,fill=relative_risk)) + 
  geom_sf(data=pts.test,  color="red", size=0.75) +
  scale_fill_viridis_b(direction=-1, option="D")
```


# CPM

## Fit Models

### setup

```{r}
load(paste0(cleandat.wd, "model_data/sa_quad.Rds"))
load(paste0(cleandat.wd, "model_data/im_list.Rds")) # im.list
```

### prepare data

```{r}
load(paste0(cleandat.wd, "model_data/pop_dat_scale.Rds")) # loads as pop.dat.scale
source(paste0(fun.path, "tbl_to_im.R")) # source function to convert tbl to im object
vars.to.im <- c(colnames(pop.dat.scale)[c(5:8, 11:23)]) # subset variable names to convert to im objects
im.list <- tbl_to_im(pop.dat.scale, vars.to.im, st_crs(3071), tmp.path = paste0(this.wd, "tmp/"), DEBUG=F) # convert those variables to im.list
save(im.list, file=paste0(cleandat.wd, "model_data/im_list.Rds")) # save im list
```

### fit null models

Only $H_{0A}$ : spatial dependence and oak wilt population

```{r}
load(paste0(cleandat.wd, "model_data/sa_quad.Rds"))
load(paste0(cleandat.wd, "model_data/im_list.Rds"))
source(paste0(fun.path, "create_null_cpm.R"))

create_null_cpm(
  method_str="clik2", 
  improve_str="wclik1", 
  cluster_type="Cauchy", 
  save_name="c2wc1")

create_null_cpm(
  method_str="clik2", 
  improve_str="wclik1", 
  cluster_type="VarGamma", 
  save_name="c2wc1")

create_null_cpm(
  method_str="palm", 
  improve_str="quasi", 
  cluster_type="Cauchy", 
  save_name="qp")

create_null_cpm(
  method_str="palm", 
  improve_str="quasi", 
  cluster_type="VarGamma", 
  save_name="qp")
```

### fit alternative models

```{r}
load(paste0(cleandat.wd, "model_data/sa_quad.Rds"))
load(paste0(cleandat.wd, "model_data/im_list.Rds"))
source(paste0(fun.path, "create_alt_cpm.R"))
source(paste0(fun.path, "create_alt_cpm_gam.R"))

create_alt_cpm(
  var_str = "sa.quad ~ surv_bool + oak_prob + bd + ch_net_dist + hs + elev + slope", 
  method_str = "clik2", 
  improve_str="wclik1", 
  cluster_type="Cauchy", 
  save_name = "c2wc1")

create_alt_cpm(
  var_str = "sa.quad ~ surv_bool + oak_prob + bd + ch_net_dist + hs + elev + slope", 
  method_str = "clik2", 
  improve_str="wclik1", 
  cluster_type="VarGamma", 
  save_name = "c2wc1"
)

create_alt_cpm(
  var_str = "sa.quad ~ surv_bool + oak_prob + bd + ch_net_dist + hs + elev + slope", 
  method_str = "palm", 
  improve_str="quasi", 
  cluster_type="Cauchy", 
  save_name = "qp")

create_alt_cpm(
  var_str = "sa.quad ~ surv_bool + oak_prob + bd + ch_net_dist + hs + elev + slope", 
  method_str = "palm", 
  improve_str="quasi", 
  cluster_type="VarGamma", 
  save_name = "qp"
)

create_alt_cpm_gam(
  var_str = "sa.quad ~ surv_bool + oak_prob + bd + ch_net_dist + hs + elev + slope", 
  method_str = "clik2", 
  cluster_type="Cauchy",
  gam_bool=T,
  save_name = "c2_gam"
)

# all vars: "full.ppp ~ surv_bool + pine_bool + oak_prob + bd + ph + sand + ch_net_dist + hs + elev + slope + aspect + plan_curv"
```

### Fit training models

```{r}
load(paste0(cleandat.wd, "model_data/train_quad.Rds")) # train.quad
load(paste0(cleandat.wd, "model_data/im_list.Rds")) # im.list
source(paste0(fun.path, "create_train_cpm.R"))
source(paste0(fun.path, "create_train_cpm_gam.R"))

create_train_cpm(
  var_str="train.quad ~ surv_bool + oak_prob", 
  method_str="clik2", 
  improve_str="wclik1", 
  cluster_type="Cauchy",
  hyp_str = "h0A",
  save_str = "c2wc1")

create_train_cpm(
  var_str="train.quad ~ surv_bool + oak_prob", 
  method_str="palm", 
  improve_str="quasi", 
  cluster_type="Cauchy",
  hyp_str = "h0A",
  save_str = "qp")

create_train_cpm(
  var_str = "train.quad ~ surv_bool + oak_prob + bd + ch_net_dist + hs + elev + slope",
  method_str="clik2", 
  improve_str="wclik1", 
  cluster_type="Cauchy",
  hyp_str = "hA",
  save_str = "c2wc1"
)

create_train_cpm(
  var_str = "train.quad ~ surv_bool + oak_prob + bd + ch_net_dist + hs + elev + slope",
  method_str="palm", 
  improve_str="quasi", 
  cluster_type="Cauchy",
  hyp_str = "hA",
  save_str = "qp"
)

create_train_cpm_gam(
  var_str = "train.quad ~ surv_bool + oak_prob + bd + ch_net_dist + hs + elev + slope",
  method_str="clik2", 
  gam_bool = T, 
  cluster_type="Cauchy",
  hyp_str = "hA",
  save_str = "c2_gam"
)

```


## Compare null to alternative model

```{r}

compare_cpm <- function(cpm_ha, cpm_h0, DEBUG=F) {
  delta.dof <- as.integer(attr(logLik(cpm_ha), "df")) - as.integer(attr(logLik(cpm_h0), "df"))
  if(DEBUG) {cat(paste0("\n\n\ndelta.dof = ", as.integer(delta.dof)) )}
  t.stat <- 2*(as.numeric(logLik(cpm_ha)) - as.numeric(logLik(cpm_h0)))
  if(DEBUG) {cat(paste0("\nt-stat = ", t.stat))}
  p.val <- pchisq(t.stat, df=delta.dof, lower.tail=F)
  if(DEBUG) cat(paste0("\np = ", p.val))
  return(p.val)
}

# 1
cpm.path <- paste0(mod.res.path, "cpm/")
load(paste0(cpm.path, "cpm_h0A_Cauchy_c2wc1.Rds"))
load(paste0(cpm.path, "cpm_hA_Cauchy_c2wc1.Rds"))
compare_cpm(cpm_hA_Cauchy_c2wc1, cpm_h0A_Cauchy_c2wc1, DEBUG=T)

# 1
load(paste0(cpm.path, "cpm_hA_Cauchy_qp.Rds"))
load(paste0(cpm.path, "cpm_h0A_Cauchy_qp.Rds"))
compare_cpm(cpm_hA_Cauchy_qp, cpm_h0A_Cauchy_qp, DEBUG=T)

AIC(cpm_h0A_Cauchy_qp) - AIC(cpm_hA_Cauchy_qp)

# 2
load(paste0(cpm.path, "cpm_hA_Cauchy_qp.Rds"))
load(paste0(cpm.path, "cpm_h0A_Cauchy_qp.Rds"))
compare_cpm(cpm_hA_Cauchy_qp, cpm_h0A_Cauchy_qp, DEBUG=T)

# 3
load(paste0(cpm.path, "cpm_hA_VarGamma_c2wc1.Rds"))
load(paste0(cpm.path, "cpm_h0A_VarGamma_c2wc1.Rds"))
compare_cpm(cpm_hA_VarGamma_c2wc1, cpm_h0A_VarGamma_c2wc1, DEBUG=T)

# 4
load(paste0(cpm.path, "cpm_hA_VarGamma_qp.Rds"))
load(paste0(cpm.path, "cpm_h0A_VarGamma_qp.Rds"))
compare_cpm(cpm_hA_VarGamma_qp, cpm_h0A_VarGamma_qp, DEBUG=T)

```

## Visualize

```{r}

load(paste0(mod.res.path,"cpm/cpm_hA_Cauchy_c2_gam.Rds"))
pred <- predict(cpm_hA_Cauchy_c2_gam) %>% as_tibble()

pred.plot <- pred %>% 
  filter(value <= 3e-05) %>% 
  mutate(relative_risk = 100*((value-min(value))/(max(value)-min(value))) )

p1 <- ggplot() +
  geom_tile(data=pred.plot, aes(x=x,y=y, fill=relative_risk)) +
  geom_sf(data=pts.test, color="red") +
  scale_fill_viridis_b(direction=-1, option="D")
```







