---
title: "model_compare"
author: "Caleb Stevens cstevens5@wisc.edu"
output: html_document
---

```{r setup, include=FALSE}
## SET PATH
this.wd <- "E:/wilt/"
##

library(tidyverse)
library(sf)
library(assertthat)
library(mgcv)
library(gridExtra)
library(spatstat)

rawdat.wd <- paste0(this.wd, "raw_data/")
cleandat.wd <- paste0(this.wd, "clean_data/")
mod.res.path <- paste0(this.wd, "models/model_obj/")
fun.path <- paste0(this.wd, "code/functions/")

```

# AIC

### Setup

```{r}
load_mod <- function(mod_str, cpm_bool) {
  if(cpm_bool) {
    return(load(paste0(mod.res.path, "cpm/", mod_str, ".Rds"), envir=globalenv()))
  } else {
    return(load(paste0(mod.res.path, mod_str, ".Rds"), envir=globalenv()))
  }
}
load_mod_select <- function() {
  cat(list.files(path = mod.res.path), sep="\n")
  mod_str <- readline(prompt="\n\nModel name : \n")
  return(load(paste0(mod.res.path, mod_str, ".Rds"), envir=globalenv()))
}
```

### GAM

```{r}
load_mod("mod_h0A_reml", F)
load_mod("mod_h0B_reml", F)
load_mod("mod_alt_reml_subs", F)
a.h01 <- AIC(mod.h0A.reml)
a.h02 <- AIC(mod.h0B.reml)
a.hA <- AIC(mod.alt.reml.subs)

cat(paste0("\nDelta AIC 1 : " , a.h01 - a.hA))
cat(paste0("\nDelta AIC 2 : " , a.h02 - a.hA))
```
### CPM

```{r}
load_mod("cpm_h0A_Cauchy_qp", T)
load_mod("cpm_hA_Cauchy_qp", T)
a.h01 <- AIC(cpm_h0A_Cauchy_qp)
a.hA <- AIC(cpm_hA_Cauchy_qp)
cat(paste0("\nDelta AIC 3 : " , a.h01 - a.hA))
```


# Classification Metrics

### Setup

```{r}
load(paste0(cleandat.wd, "model_data/oak_dat_scale.Rds"))

source(paste0(fun.path, "map_wilt.R"))
source(paste0(fun.path, "prec_rec_fun.R"))
source(paste0(fun.path, "compare_models.R"))

load_mod <- function(mod_str, cpm_bool) {
  if(cpm_bool) {
    return(load(paste0(mod.res.path, "cpm/", mod_str, ".Rds"), envir=globalenv()))
  } else {
    return(load(paste0(mod.res.path, mod_str, ".Rds"), envir=globalenv()))
  }
}

load_mod_select <- function() {
  cat(list.files(path = mod.res.path), sep="\n")
  mod_str <- readline(prompt="\n\nModel name : \n")
  return(load(paste0(mod.res.path, mod_str, ".Rds"), envir=globalenv()))
}

mod.list.path <- paste0(this.wd, "models/model_lists/")
mod.comp.path <- paste0(this.wd, "models/model_comp/")
load(paste0(mod.comp.path, "test_pts_oak_tbl.Rds"))
```

### Create model testing reference tibbles

```{r}
# # all 108 test pts are in study area, but how many are in oak
# load(paste0(cleandat.wd, "model_data/pop_dat_scale.Rds"))
# 
# pop.dat.scale %>% 
#   filter(wilt_test >= 1) %>% 
#   filter(oak == 1) %>% 
#   summarise(ct = n())
# # 81 are within oaks
# 
# mod.comp.path <- paste0(this.wd, "models/model_comp/")
# 
# # to filter down to just the test.pts that are within oak -
# test.pts.pop.tbl <- pop.dat.scale %>% 
#   filter(wilt_test >= 1) %>% 
#   select(x, y, wilt_test) %>% 
#   mutate(x = (x*attr(x, "scaled:scale")) + attr(x, "scaled:center"), # unscale x
#                y = (y*attr(y, "scaled:scale")) + attr(y, "scaled:center")) %>%  # unscale y
#   rename(wilt = wilt_test)
# 
# save(test.pts.pop.tbl, file=paste0(mod.comp.path, "test_pts_pop_tbl.Rds") )
# 
# test.pts.oak.tbl <- pop.dat.scale %>% 
#   filter(wilt_test >= 1) %>% 
#   filter(oak == 1) %>% 
#   select(x, y, wilt_test) %>% 
#   mutate(x = (x*attr(x, "scaled:scale")) + attr(x, "scaled:center"), # unscale x
#                y = (y*attr(y, "scaled:scale")) + attr(y, "scaled:center")) %>%  # unscale y
#   rename(wilt = wilt_test)
# 
# save(test.pts.oak.tbl, file=paste0(mod.comp.path, "test_pts_oak_tbl.Rds"))
```

### Compare performance between GAM models as predictive classifiers

```{r}
# load(paste0(mod.comp.path, "test_pts_oak_tbl.Rds"))

# load_mod("mod_train_alt_reml") # mod.train.alt.reml
# load_mod("mod_train_alt_gacv", F) # mod.train.alt.gacv
# load_mod("mod_train_h0A_reml") # mod.train.h0A.reml
# load_mod("mod_train_h0A_gacv", F) # mod.train.h0A.gacv
# load_mod("mod_train_h0B_reml") # mod.train.h0B.reml
# load_mod("mod_train_h0B_gacv", F) # mod.train.h0B.gacv

# hp3.list <- list(ha_gacv = mod.train.alt.gacv, h0A_gacv = mod.train.h0A.gacv, h0B_gacv = mod.train.h0B.gacv)
# hp3.list <- list(ha_reml = mod.train.alt.reml, ha_gacv = mod.train.alt.gacv, h0A_reml = mod.train.h0A.reml, h0A_gacv = mod.train.h0A.gacv, h0B_reml = mod.train.h0B.reml, h0B_gacv = mod.train.h0B.gacv)
# save(hp3.list, file=paste0(mod.list.path, "hp3_list.Rds"))

# load HP3 models
#load(paste0(mod.list.path, "hp3_list.Rds")) # hp3.list
# mod.comp.tbl <- compare_models(hp3.list, test.pts.oak.tbl, hp3_bool_v=rep(1,3), hp3_scale_v=rep(0,3), DEBUG_SUB=F, DEBUG=T)
# save(mod.comp.tbl, file=paste0(this.wd, "models/model_comp/hp3_gam_comp_tbl.Rds"))
load(paste0(mod.comp.path, "hp3_gam_comp_tbl.Rds")) # loads as mod.comp.tbl

plot.comp.tbl <- mod.comp.tbl %>% 
  pivot_wider(
    id_cols = c(mod_name, thresh), 
    names_from=measure, 
    values_from = val) %>% 
  mutate(area_surveyed_scale = area_surveyed*0.54)

p1 <- plot.comp.tbl %>% 
    ggplot(aes(x=area_surveyed_scale*100, y=recall*100, color=mod_name)) +
    geom_line(size=0.7) +  
    geom_point(aes(shape=mod_name), size=4) +
    
    scale_shape_manual(
      name=bquote(HP^3 ~ Model),
      values = c("h0A_gacv"= 1, "h0B_gacv"= 3, "ha_gacv"= 2),
      labels= c(bquote(H["01"]), bquote(H["02"]), bquote(H[A]))
    ) +
  
    scale_color_viridis_d(
      name=bquote(HP^3 ~ Model), 
      labels= c(bquote(H["01"]), bquote(H["02"]), bquote(H[A]) )
    ) +
    
    labs(x="Area Surveyed (%)", y="Recall (%)") +
  
    theme(legend.key.size = unit(2, "cm"),
          legend.text = element_text(size=20),
          legend.title = element_text(size=24),
          legend.position = c(0.9, 0.4),
          axis.text = element_text(size=18),
          axis.title = element_text(size=24))

```

### Compare between CPMs

```{r}
#load_mod("cpm_train_h0A_Cauchy_qp", T)
#load_mod("cpm_train_hA_Cauchy_qp", T)
# load_mod("cpm_train_h0A_Cauchy_c2wc1", T)
# load_mod("cpm_train_hA_Cauchy_c2_gam", T)
# load_mod("cpm_train_hA_Cauchy_c2wc1", T)

# load(paste0(cleandat.wd, "model_data/sa_quad.Rds"))
# test.pts <- st_read(paste0(cleandat.wd, "sa_points/sa_points.shp")) %>% filter(!(is.na(year)), year >= 2019)
# 
# # cpm.list <- list(
# #   #h0A_qp = cpm_train_h0A_Cauchy_qp,
# #   #hA_qp = cpm_train_hA_Cauchy_qp,
# #   h0A_c2wc1 = cpm_train_h0A_Cauchy_c2wc1,
# #   hA_c2_gam = cpm_train_hA_Cauchy_c2_gam,
# #   hA_c2wc1 = cpm_train_hA_Cauchy_c2wc1)
# 
# #save(cpm.list, file=paste0(mod.list.path,"cpm_list.Rds"))
# load(paste0(mod.list.path, "cpm_list.Rds"))
# 
# # load(paste0(mod.list.path,"cpm_list.Rds")) # cpm.list
# # load(paste0(mod.comp.path, "test_pts_pop_tbl.Rds")) # test.pts.pop.tbl
# 
# cpm.comp.tbl <- compare_models(cpm.list, test.pts.pop.tbl, hp3_bool_v=rep(0,3), hp3_scale_v=rep(0,3), DEBUG_SUB=T, DEBUG=T)
# save(cpm.comp.tbl, file=paste0(this.wd, "models/model_comp/cpm_comp_tbl.Rds"))
# 

load(paste0(mod.comp.path, "cpm_comp_tbl.Rds"))

plot.comp.tbl <- cpm.comp.tbl %>% 
  group_by(measure) %>% 
  mutate(val_scale = val/max(val)) %>% 
  ungroup() %>% 
  pivot_wider(id_cols=c(mod_name, scale_thresh) , names_from = measure, values_from = val_scale) #%>% 
  #pivot_longer(cols=recall:rel_eff,names_to = "measure", values_to="val_scale")

# p0 <- plot.comp.tbl %>% 
#   ggplot(aes(x=area_surveyed*100, y=recall*100, color=mod_name)) + 
#     geom_line(size=1.1) + 
#     labs(x="% Area Surveyed", y="Recall (% Cases Predicted)") +
#     theme(title = element_text(size=32, hjust=0.5), axis.text = element_text(size=8)) +
#   scale_color_discrete(name="Model Name") +
#   theme(legend.key.size = unit(2, "cm"),
#           legend.text = element_text(size=20))


p1 <- plot.comp.tbl %>% 
  ggplot(aes(x=area_surveyed*100, y=recall*100, color=mod_name)) + 
    geom_line(size=0.7) +
    geom_point(aes(shape=mod_name), size=4) +
      
  scale_shape_manual(
    name="CPM Model",
    values = c("h0A_c2wc1"= 1, "hA_c2_gam"= 3, "hA_c2wc1"= 2),
    labels=c( bquote(H["01"]), bquote(H[A] ~ GAM), bquote(H[A] ~ Linear) )
  ) +
  
  scale_color_viridis_d(
    name="CPM Model" , 
    option="E",
    labels=c( bquote(H["01"]), bquote(H[A] ~ GAM), bquote(H[A] ~ Linear) ) ) +
  
  labs(x="Area Surveyed (%)", y="Recall (%)") +
  
  theme(legend.key.size = unit(2, "cm"),
          legend.text = element_text(size=20),
          legend.title = element_text(size=24),
          legend.position = c(0.8, 0.4),
          axis.text = element_text(size=18),
          axis.title = element_text(size=24))
  
  

```

### Compare CPM to HP3-GAM

```{r}
load_mod("mod_train_alt_gacv", F)
load_mod("cpm_train_hA_Cauchy_c2wc1", T)
load(paste0(mod.comp.path, "test_pts_oak_tbl.Rds"))
load(paste0(cleandat.wd, "model_data/sa_quad.Rds"))
test.pts <- st_read(paste0(cleandat.wd, "sa_points/sa_points.shp")) %>% filter(!(is.na(year)), year >= 2019)

mod.comp.list <- list(
   hp3_gacv = mod.train.alt.gacv,
   cpm_c2wc1 = cpm_train_hA_Cauchy_c2wc1)

across.comp.tbl <- compare_models(mod.comp.list, test.pts.oak.tbl,  hp3_bool_v=c(1,0), hp3_scale_v=c(1,0), DEBUG_SUB=T, DEBUG=T)

save(across.comp.tbl, file=paste0(mod.comp.path, "across_comp_tbl.Rds"))
load(paste0(mod.comp.path, "across_comp_tbl.Rds"))

plot.dat <- across.comp.tbl %>% 
  select(-thresh) %>% 
  filter(measure %in% c("area_surveyed", "recall")) %>% 
  pivot_wider(id_cols = c(scale_thresh, mod_name), names_from = measure, values_from = val)

# remove half of the hp3_gacv indices to prevent over-plotting
cpm.ind <- which(plot.dat$mod_name == "cpm_c2wc1")
hp3.ind <- which(plot.dat$mod_name == "hp3_gacv")
ind.tmp <- hp3.ind[seq(from=1, to=300, by=2)]
plot.dat.subs <- plot.dat[c(cpm.ind,ind.tmp), ]

p1 <- plot.dat.subs %>% 
  
  ggplot(aes(x=area_surveyed*100, y=recall*100, color=mod_name)) +
    geom_line(size=0.7) +
    geom_point(aes(shape=mod_name), size=4) +
    scale_shape_manual(
      name=bquote(H[A] ~ Model),
      values = c("cpm_c2wc1"= 1, "hp3_gacv"= 2),
      labels=c(  bquote(CPM), bquote(HP^3~-GAM))
    ) +
  scale_color_viridis_d(
    name=bquote(H[A] ~ Model),
    begin=0,
    end=0.7,
    option="E",
    labels=c( bquote(CPM), bquote(HP^3~-GAM)) 
  ) +
  labs(x="Area Surveyed (%)", y="Recall (%)") +
  
  theme(legend.key.size = unit(2, "cm"),
          legend.text = element_text(size=20),
          legend.title = element_text(size=24),
          legend.position = c(0.8, 0.4),
          axis.text = element_text(size=18),
          axis.title = element_text(size=24))
   

```

# Hazard Maps

### setup

```{r}
load_mod <- function(mod_str, cpm_bool) {
  if(cpm_bool) {
    return(load(paste0(mod.res.path, "cpm/", mod_str, ".Rds"), envir=globalenv()))
  } else {
    return(load(paste0(mod.res.path, mod_str, ".Rds"), envir=globalenv()))
  }
}

plot.path <- paste0(this.wd, "figures/hazard/")

```

### HP3-GAM

Make predictions and export them

```{r}
# load(paste0(cleandat.wd, "model_data/oak_dat_scale.Rds"))
# load(paste0(cleandat.wd, "merge/join_cts_tbl.Rds"))
# pred.dat <- join.cts.tbl %>% 
#   filter(sa==1, oak==1, priv==0, water==0) %>% 
#   select(x, y, wilt)
# save(pred.dat, file=paste0(cleandat.wd, "pred_dat/pred_dat.Rds") )

load(paste0(cleandat.wd, "pred_dat/pred_dat.Rds"))

load_mod("mod_alt_gacv_subs", F)

p <- predict(mod.alt.gacv.subs, type="response", se.fit=T)

m.pred.tbl <- tibble(
    pred = p$fit, 
    se=p$se.fit, 
    x = pred.dat$x, 
    y = pred.dat$y, 
    wilt= pred.dat$wilt) %>%  
  mutate(pred = pred/max(pred),
         confidence = 1 - se/max(se) )

# p1 <- m.pred.tbl %>% 
#   ggplot() +
#     geom_point(aes(x=x,y=y, color=pred, alpha=1-confidence))

# how many points greater than 0.2?
m.pred.tbl %>% 
  filter(pred > 0.4) %>%
  summarise(count = n())

# convert to points
pred.pts <- m.pred.tbl %>% 
  filter(pred > 0.4) %>% 
  select(x, y, pred, se, confidence) %>% 
  st_as_sf(coords=c("x", "y"),
           crs=st_crs(3071))

write.path <- "D:/Backed Up/Desktop/Wilt/"

st_write(pred.pts, dsn=paste0(write.path, "combined_pred_pts/combined_pred_pts.shp") )



```

**Generate plots - might need work**

```{r}
wilt.pts <- m.pred.tbl %>% 
  filter(wilt==1) %>% 
  select(x, y, wilt) %>% 
  mutate(wilt = "2004-2021")

p1 <- m.pred.tbl %>% 
  rename(hazard=pred) %>% 
  ggplot() + 
    geom_tile(aes(x=x,y=y,fill=hazard)) +
    geom_point(data=wilt.pts, aes(x=x,y=y, shape=wilt), size=3, alpha=0.8, color="black") +
    scale_fill_viridis_c(option="B", direction=-1) +
    scale_shape_manual(values=c("2004-2021" = 4)) +
    theme_bw() +
    guides(fill = guide_colorbar(title="Hazard",order=1),
          shape = guide_legend(title="Wilt", order=2)) +
    theme(legend.key.size = unit(1, "cm"),
          legend.text = element_text(size=20),
          legend.title = element_text(size=24),
          legend.position = c(0.85, 0.3),
          axis.text = element_text(size=18),
          axis.title = element_text(size=24))

# ggsave(filename = "hp3_haz.tif",
#        plot = p1,
#        device="tiff",
#        path = plot.path,
#        width = 900,
#        height = 900,
#        units = "px",
#        dpi = 600)

p2 <- m.pred.tbl %>% 
  ggplot() + 
    geom_tile(aes(x=x,y=y,fill=se)) +
    geom_point(data=wilt.pts, aes(x=x,y=y, shape=wilt), size=3, alpha=0.8, color="black") +
    scale_fill_viridis_c(option="D", direction=-1) +
    scale_shape_manual(values=c("2004-2021" = 4)) +
    theme_bw() +
    guides(fill = guide_colorbar(title="Uncertainty", order=1),
          shape = guide_legend(title="Wilt", order=2) ) +
    theme(legend.key.size = unit(1, "cm"),
          legend.text = element_text(size=20),
          legend.title = element_text(size=24),
          legend.position = c(0.85, 0.3),
          axis.text = element_text(size=18),
          axis.title = element_text(size=24))

# ggsave(filename = "hp3_se.tif",
#        plot = p2,
#        device = "tiff",
#        path = plot.path,
#        scale = 1,
#        width = 900,
#        height = 900,
#        units = "px",
#        dpi = 600)
```

Generate predictions from habitat model and export

```{r}
load_mod("mod_h0B_gacv_subs", F)
p <- predict(mod.h0B.gacv.subs, type="response", se.fit=T)
m.pred.tbl <- tibble(
    pred = p$fit, 
    se = p$se.fit, 
    x = pred.dat$x, 
    y = pred.dat$y, 
    wilt= pred.dat$wilt) %>% 
  mutate(pred = pred/max(pred),
         confidence = 1 - se/max(se))
m.pred.tbl %>% 
  filter(pred > 0.33) %>%
  summarise(count = n())
pred.subs.pts <- m.pred.tbl %>% 
  filter(pred > 0.33) %>% 
  select(x, y, pred, se, confidence) %>% 
  st_as_sf(coords=c("x", "y"),
           crs=st_crs(3071))
st_write(pred.subs.pts, dsn=paste0(write.path, "habitat_pred_pts/habitat_pred_pts.shp") )
```



### CPM

```{r}
load_mod("cpm_hA_Cauchy_c2wc1", T)
load(paste0(cleandat.wd, "model_data/sa_quad.Rds"))
load(paste0(cleandat.wd, "model_data/im_list.Rds"))

sa.pts <- st_read(paste0(cleandat.wd, "sa_points/sa_points.shp")) %>% 
  mutate(year = "2004-2021")

sa.pts.tbl <- do.call(rbind, st_geometry(sa.pts)) %>% 
  as_tibble() %>% 
  setNames(c("x", "y")) %>% 
  add_column(year = "2004-2021")

p <- predict.kppm(cpm_hA_Cauchy_c2wc1, locations=sa.quad$dummy, covariates=im.list, se=T)

cpm.pred.tbl <- tibble(
    pred = p$estimate, 
    se=p$se, 
    x=sa.quad$dummy$x, 
    y=sa.quad$dummy$y) %>% 
  mutate(pred = pred/max(pred, na.rm=T))

p1 <- cpm.pred.tbl %>% 
  ggplot() +
    geom_tile(aes(x=x,y=y,fill=pred)) +
    geom_point(data=sa.pts.tbl, aes(x=x, y=y, shape=year), size=3, alpha=0.8, color="black") +
    scale_fill_viridis_c(option="B", direction=-1) +
    scale_shape_manual(values=c("2004-2021" = 4)) +
    theme_bw() +
    guides(fill = guide_colorbar(title="Hazard",order=1),
          shape = guide_legend(title="Wilt", order=2)) +
    theme(legend.key.size = unit(1, "cm"),
          legend.text = element_text(size=20),
          legend.title = element_text(size=24),
          legend.position = c(0.85, 0.3),
          axis.text = element_text(size=18),
          axis.title = element_text(size=24))

p2 <- cpm.pred.tbl %>% 
  ggplot() +
    geom_tile(aes(x=x,y=y,fill=se)) +
    geom_point(data=sa.pts.tbl, aes(x=x, y=y, shape=year), size=3, alpha=0.8, color="black") +
    scale_fill_viridis_c(option="D", direction=-1) +
    scale_shape_manual(values=c("2004-2021" = 4)) +
    theme_bw() + 
    guides(fill = guide_colorbar(title="Effect SE",order=1),
          shape = guide_legend(title="Wilt", order=2)) +
    theme(legend.key.size = unit(1, "cm"),
          legend.text = element_text(size=20),
          legend.title = element_text(size=24),
          legend.position = c(0.85, 0.3),
          axis.text = element_text(size=18),
          axis.title = element_text(size=24))

```



# Model Diagnostics

## residual analysis

### setup

```{r}
load_mod <- function(mod_str, is_cpm) {
  if(is_cpm) {
    return(load(paste0(mod.res.path, "cpm/", mod_str, ".Rds"), envir=globalenv()))
  } else {
    return(load(paste0(mod.res.path, mod_str, ".Rds"), envir=globalenv()))
  }
}
source(paste0(fun.path, "resid_fun.R"))
source(paste0(fun.path, "resid_plot_fun.R"))

load(paste0(cleandat.wd, "model_data/sa_quad.Rds"))
sa.pts <- st_read(dsn=paste0(cleandat.wd, "sa_points/sa_points.shp"))
```

### GAM

```{r}
load_mod("mod_alt_gacv_subs", is_cpm=F) # mod.alt.gacv.subs
load(paste0(cleandat.wd, "model_data/oak_dat_scale.Rds")) # oak.dat.scale

res.tmp <- resid_fun(
  mod=mod.alt.gacv.subs, 
  is_kppm=0, 
  dat_tbl=oak.dat.scale, 
  pts_shp=sa.pts, 
  p_slice=20, 
  dbg=T)

resid_plot_fun(res.tmp)

rm(res.tmp)
```

### CPM

```{r}
# load
load_mod("cpm_hA_Cauchy_c2wc1", is_cpm=T) # cpm_hA_Cauchy_c2_gam
load(paste0(cleandat.wd, "model_data/pop_dat_scale.Rds")) # pop.dat.scale
# clean
pop.dat.sa <- pop.dat.scale %>% 
  mutate(x= (x*attr(x, "scaled:scale"))+attr(x, "scaled:center"),
         y = (y*attr(y, "scaled:scale"))+attr(y, "scaled:center")) %>% 
  select(x, y, oak_prob, elev, plan_curv, ch_net_dist, bd, ph, hs, sand)
# construct residual plots
res.tmp <- resid_fun(
  mod=cpm_hA_Cauchy_c2wc1, 
  is_kppm=1, 
  dat_tbl=pop.dat.sa, 
  pts_shp=sa.pts, 
  p_slice=20, 
  dbg=T)
# plot & arrange
resid_plot_fun(res.tmp)
```
